% the acknowledgments section
\begin{doublespace}

There are an incredible number of people who helped me along the path to earning a doctorate. I wish I could recognize and thank everyone here, but the list of people who have supported me personally and professionally throughout my life is impossibly long. What follows is certainly an incomplete list; to everyone else, know that I appreciate the presence and impact you have had in my life.

First and foremost, I would like to thank my research advisor, Gary Feldman, for your advice and support throughout my time at Harvard. Your input, understanding, and guidance helped me grow and find success. Thank you to my other committee members, Melissa Franklin and Howard Georgi. I am grateful for your physics advice and for the positive environment you helped to shape in the physics department. I would also like to thank my original graduate academic advisor, Matthew Schwartz, for your support and encouragement through my initial years as a graduate student.

Thank you to the \nova~collaboration, whose work produced the wonderful experiment I was able to be part of. In particular, thank you to the NC/Sterile group conveners, Alex Sousa and Louise Suter, for your guidance and support, especially during the analysis push. Thank you to the other contributors to the NC disappearance analysis, Adam Aurisano, Aaron Markowitz, and Shaokai Yang. The analysis had one other major contributor who I believe deserves special recognition, Gavin Davies. Thank you for your part in the analysis, being a friend, allowing me to vent my frustrations, and for all your help before and during my work on the NC disappearance analysis; I owe you a great debt of gratitude. Thank you also to my fellow collaborators Michael Baird, Xuebing Bu, Joseph Lozier, Evan Niner, Dan Pershey, Fernanda Psihas, Justin Vasel, for your friendship.

The post docs I worked with in the Harvard neutrino group were truly instrumental to my graduate work. Thank you to Raphael Schroeter and Ruth Toner; you both truly cared about my success and I was able to achieve it with your guidance and support. I will call out Alex Sousa again here-thank you for your part in providing a positive research experience (and for sharing your fantastic knowledge of restaurants in Naperville) while we overlapped at Harvard.

There are several other people in the Harvard Department of Physics who deserve recognition. Thank you to Jacob Barandes for your advice throughout the graduate process. Thank you, Lisa Cacciabaudo and Carol Davis, for your open doors and tasty treats. Many graduate students shared late nights with me as we worked through problem sets. Thank you specifically to Mike Goldman who worked with me through several courses and has remained a good friend. Thank you to my current and former Palfrey house denizens, Stephan Chan, Tomo Lazovich, Chris Rogan, Emma Tolley, and Ann Wang. I will fondly remember our Taco Tuesdays, especially the ones on the Felipe's roof deck.

I wish to thank two people from my time as an undergraduate at CWRU, my advisor, Glenn Starkman, and my senior project advisor, Robert ``Doc'' Brown. Both of you encouraged me personally and professionally, and I owe much of my success to the mentorship, opportunities, and encouragement you provided me.

Last, but certainly not least, is my family. Thank you to my mother and father, Shelly and Kris Kafka, for all of your encouragement through my entire life. You always believed in me and pushed me to achieve greatness. Both of you have had tremendous influence on how I approach the world, and I cannot imagine how I would have accomplished as much without your impact and love. Thank you to my brother Wesley for inspiring me to be the best version of myself possible. I want always to be a good role model for you, and it brings me endless pride to see what you have accomplished so far. To my adopted brothers, Scott and Cody, thank you for always welcoming me into a home away from home. I am now a fully fledged wizard. Thank you to Karen and Glenn Berry, my mother-in-law and father-in-law, for always being so welcoming and supportive. Finally, I extend a most heartfelt thanks to my wife, Kristen. You provided comfort on my worst days and amplified the triumphs of my best days, always having faith in me. Your constant love and support gave me the strength to complete my degree.

\end{doublespace}